---
title: "Reproducable research - Assignment week 2"
author: "YOU"
date: "3/28/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = FALSE}

rm(list = ls())

library(knitr)
library(ggplot2)
library(RColorBrewer)
library(grid)
library(dplyr)

```


## Data description

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day and include the number of steps taken in 5 minute intervals each day.


The variables included in this dataset are:

1. steps: Number of steps taking in a 5-minute interval

2. date: The date on which the measurement was taken

3. interval: Identifier for the 5-minute interval


For this assignment I loaded the following R packages:

- Knitr
- ggplot2
- RColorBrewer
- grid
- dplyr


#### Loading the data

```{r}
  dataFile <- read.csv("activity.csv", header = TRUE)

  dataFile$date <- as.Date(dataFile$date)
```



## Total steps by day


Calculate the total number of steps taken per day and plot it as a histogram

```{r, echo = FALSE}
  totalSteps <- aggregate(dataFile$steps, by=list(dataFile$date), FUN = sum, na.rm = TRUE)
  colnames(totalSteps) <- c("date", "stepsByDay")
  
  ggplot(totalSteps, aes(stepsByDay)) + 
    geom_histogram(color="black", fill="white", bins = 12) + 
    labs(title = "Total steps by day",
       x = "Steps per Day", 
       y = "Count")

```


Calculate and report the mean and median of the total number of steps taken by day

Median: 
```{r, echo = TRUE}
  median(totalSteps$stepsByDay, na.rm = TRUE)
```

Mean: 
```{r, echo = TRUE}
  mean(totalSteps$stepsByDay, na.rm = TRUE)
```



## Average steps by interval 

Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r, echo = TRUE}
  averageStepsByInterval <- aggregate(dataFile$steps, by=list(dataFile$interval), FUN = mean, na.rm = TRUE)
  colnames(averageStepsByInterval) <- c("interval", "AVGSteps")

  ggplot(averageStepsByInterval, aes(interval, AVGSteps)) + 
      geom_line(color = "blue", size = 1) + 
      labs(title = "Average steps by 5-minute interval",
          x = "Interval", 
          y = "Average Steps")

```


5-minute interval that (on average across all days) contains the maximum number of steps: 
```{r, echo = TRUE} 
  averageStepsByInterval$interval[which.max(averageStepsByInterval$AVGSteps)]
```


## Missing Values


Total number of missing values in the dataset:
```{r, echo = TRUE}
  sum(is.na(dataFile$steps))
```


Filling in missing values: 

1. Calculate the average steps by day

2. Impute the average steps by day to the fields with missing values

```{r, echo = TRUE}
  fieldsNA <- averageStepsByInterval$AVGSteps[match(dataFile$interval, averageStepsByInterval$interval)]
  
  dataFileWithoutNA <- transform(dataFile, steps = ifelse(is.na(dataFile$steps), yes = fieldsNA, no = dataFile$steps))
```


Make a histogram of the total number of steps taken each day (dataset without missing values)

```{r, echo = TRUE}
  totalStepsWithoutNA <- aggregate(list(dataFileWithoutNA$interval, dataFileWithoutNA$steps), by=list(dataFileWithoutNA$date), FUN = sum) 
  colnames(totalStepsWithoutNA) <- c("date", "interval", "stepsByDay")

  
  ggplot(totalStepsWithoutNA, aes(stepsByDay)) + 
     geom_histogram(color="black", fill="white", bins = 12) + 
     labs(title = "Total steps by day",
         x = "Steps per Day", 
         y = "Count")
```

Calculate and report the mean and median of the total number of steps taken by day

Median: 
```{r, echo = TRUE}
  median(totalStepsWithoutNA$stepsByDay, na.rm = TRUE)
```

Mean: 
```{r, echo = TRUE}
  mean(totalStepsWithoutNA$stepsByDay, na.rm = TRUE)
```



## Weekday vs. Weekend


Make a panel plot containing a time series plot of the 5-minute interval and the average number of steps taken, averaged across all weekday or weekend days. 

```{r, echo = TRUE}
  dataFileWithoutNA$weekday <- weekdays(dataFile$date)
  dataFileWithoutNA$dayType <- ifelse(dataFileWithoutNA$weekday=='Saturday' | dataFileWithoutNA$weekday=='Sunday', 'Weekend','Weekday')
  
  averageStepsByInterval2 <- aggregate(steps~interval + dayType, dataFileWithoutNA, FUN = mean, na.rm = TRUE)
  colnames(averageStepsByInterval2) <- c("interval", "dayType", "AVGSteps")
  
  par(mfrow = c(2,1))
  ggplot(averageStepsByInterval2, aes(interval, AVGSteps, color = dayType)) +
      geom_line() +
      labs(title = "Average steps by 5-minute interval", 
           x = "Interval", 
           y = "Average steps") + 
    facet_wrap(~dayType, ncol = 1, nrow = 2 )

``` 
